The evolution of gene expression in the term placenta of viviparous mammals

### 2015-10-02 manuscript/manuscript.Rnw
add table of sequencing results and figure of read efficiency

<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
opts_chunk$set(cache=TRUE, autodep=TRUE)
options(device = function(file, width = 8, height = 7, ...) {
  cairo_pdf(tempfile(), width = width, height = height, ...)
})
library("data.table")
library("ggplot2")
library("reshape2")
library("xtable")
@

@@

<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm")))>>=
for (a in paste0("../data/",c("combined_fpkm"))) {
    load(a)
}
@

@@

The placenta is arguably the most anatomically variable organ in
mammals. Understanding gene expression patterns among placentas from
multiple species can illuminate the molecular basis of placenta
function as well as potentially enrich our knowledge of the biology of
obstetrical syndromes. We profiled the term placenta by using Illumina
RNA-Seq technology to sequence the transcriptomes of 6 species of
therian (i.e. marsupials and “placental”) mammals. We identified
significant shifts in gene expression along specific evolutionary
lineages using a model of character evolution. Genes that showed
significant elevation in expression along the human lineage
($\mathrm{FDR}\le 0.01$) were enriched in GO categories such as
angiogenesis, tube morphogenesis, cell migration, and regulation of
inflammatory response. Three of these genes (KISS1, ADAM12, GPC3)
showed exceptionally high expression in humans; these genes are
implicated in preeclampsia and may indicate that the risk of
preeclampsia recently emerged on the human lineage.

Keywords: mammal, gene expression, evolution, RNA-seq, placenta

@@

\begin{table}
  \begin{tabular}{c c c c}
    \toprule
    Species                       & Common Name        & Reference & Annotation \\
    \midrule
    \textit{Dasypus novemcinctus} & 9-Banded Armadillo & DASNOV3    & Ensembl 80 \\
    \textit{Ateles fusciceps}     & Spider monkey      & ---        & \textit{Homo sapiens} Ensembl 80 \\
    \textit{Pan paniscus}         & Bonobo             & CHIMP2.1.4 & Ensembl 80 \\
    \textit{Bos taurus}           & Cow                & UMD3.1     & Ensembl 80 \\
    \textit{Canis familiaris}     & Dog                & CanFam3.1  & Ensembl 80 \\
    \textit{Loxodonta africana}   & Elephant           & loxAfr3    & Ensembl 80 \\
    \textit{Homo sapiens}         & Human              & GRCh38     & Ensembl 80 \\
    \textit{Mus musculus}         & Mouse              & GRCm38     & Ensembl 80 \\
    \textit{Nannospalax galili}   & Blind mole rat     & ----       & \textit{Mus musculus} Ensembl 80 \\
    \textit{Monodelphis domestica} & Opossum           & BROAD05    & Ensembl 80 \\
    \textit{Spalax carmeli}       & Blind mole rat     & ----       & \textit{Mus musculus} Ensembl 80 \\
    \bottomrule
  \end{tabular}
  \caption{Reference genomes for aligned species}
  \label{tab:ref_genome_aligned_species}
\end{table}

@@

\begin{figure}
<<transcriptome.mapping.figure,fig.width=8,fig.height=8,echo=FALSE>>=
print(ggplot(star.logs[grepl("%",field),],
             aes(x=species,y=as.numeric(gsub("%","",value)),
                 color=field))
      + geom_boxplot()
      + facet_grid(field~.,scales="free")
      + ylab("Percent of Reads")
      + theme(legend.position="top",
              axis.text.x = element_text(angle = 30, hjust = 1, vjust=1))
      )
@ 
\caption{Transcript statistics for each species}
\label{fig:transcriptome_statistics}
\end{figure}

<<transcriptome.statistics>>=
@

>>>>

<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
opts_chunk$set(cache=TRUE, autodep=TRUE)
options(device = function(file, width = 8, height = 7, ...) {
  cairo_pdf(tempfile(), width = width, height = height, ...)
})
library("data.table")
library("ggplot2")
library("reshape2")
library("xtable")
library("grid")

capwords <- function(s, strict = FALSE) {
    cap <- function(s) paste(toupper(substring(s, 1, 1)),
                             {s <- substring(s, 2); if(strict) tolower(s) else s},
                             sep = "", collapse = " " )
    sapply(strsplit(s, split = " "), cap, USE.NAMES = !is.null(names(s)))
}

capfirst <- function(s, strict = FALSE) {
    paste(toupper(substring(s, 1, 1)),
          {s <- substring(s, 2); if(strict) tolower(s) else s},
          sep = "")
}

@

@@

<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm")))>>=
for (a in paste0("../data/",c("combined_fpkm"))) {
    load(a)
}
star.logs[,value:=gsub("%$","",value)]
@

@@

The placenta is arguably the most anatomically variable organ in
mammals. Understanding gene expression patterns among placentas from
multiple species can illuminate the molecular basis of placenta
function as well as potentially enrich our knowledge of the biology of
obstetrical syndromes. We profiled the term placenta by using Illumina
RNA-Seq technology to sequence the transcriptomes of
\Sexpr{star.logs[,length(unique(species))]} species of therian (i.e.
marsupials and “placental”) mammals. We identified significant shifts
in gene expression along specific evolutionary lineages using a model
of character evolution. Genes that showed significant elevation in
expression along the human lineage ($\mathrm{FDR}\le 0.01$) were
enriched in GO categories such as angiogenesis, tube morphogenesis,
cell migration, and regulation of inflammatory response. Three of
these genes (KISS1, ADAM12, GPC3) showed exceptionally high expression
in humans; these genes are implicated in preeclampsia and may indicate
that the risk of preeclampsia recently emerged on the human lineage.

\textbf{Keywords:} mammal, gene expression, evolution, RNA-seq, placenta

@@

\begin{table}
  \begin{tabular}{c c c c}
    \toprule
    Species                       & Common Name        & Reference & Read Length  \\
    \midrule
    \textit{Ateles fusciceps}     & Spider monkey      & \textit{Homo sapiens} & 100 \\
    \textit{Bos taurus}           & Cow                & UMD3.1     & 76 \\
    \textit{Canis familiaris}     & Dog                & CanFam3.1  & 151 \\
    \textit{Dasypus novemcinctus} & 9-Banded Armadillo & DASNOV3    & 101 \\
    \textit{Loxodonta africana}   & Elephant           & loxAfr3    & 76 \\
    \textit{Homo sapiens}         & Human              & GRCh38     & 76, 151 \\
    \textit{Monodelphis domestica} & Opossum           & BROAD05    & 76 \\
    \textit{Mus musculus}         & Mouse              & GRCm38     & 36,44,151 \\
    \textit{Nannospalax galili}   & Blind mole rat     & \textit{Mus musculus} & 100 \\
    \textit{Pan paniscus}         & Bonobo             & CHIMP2.1.4 & 151 \\
    \textit{Spalax carmeli}       & Blind mole rat     & \textit{Mus musculus} & 100 \\
    \bottomrule
  \end{tabular}
  \caption{Reference genomes for aligned species}
  \label{tab:ref_genome_aligned_species}
\end{table}

@@

\begin{figure}
<<transcriptome.mapping.figure,fig.width=8,fig.height=8,echo=FALSE>>=
star.logs.unmapped <- 
    star.logs[grepl("%",field) & 
                  grepl("unmapped",field) & 
                      !grepl("too many mismatches",field),]
star.logs.unmapped <- 
    star.logs.unmapped[,field:=capfirst(gsub("\\% of reads unmapped: ","",field))]
pushViewport(viewport(layout=grid.layout(ncol=1,nrow=2,heights=c(2,1))))
print(ggplot(star.logs.unmapped,
             aes(x=capfirst(species),y=as.numeric(gsub("%","",value))))
      + geom_boxplot()
      + facet_grid(field~.,scales="free")
      + ylab("Percent of Reads")
      + xlab("Species")
      + theme(legend.position="top",
              axis.text.x = element_text(angle = 30, hjust = 1, vjust=1,face="italic")),
      vp=viewport(layout.pos.col=1,layout.pos.row=1)
      )
pushViewport(viewport(layout.pos.col=1,layout.pos.row=1))
grid.text("A",
          y=unit(1,"npc")-unit(0.1,"cm"),
          x=unit(0,"npc")+unit(0.1,"cm"),
          gp=gpar(fontsize=24),
          just=c("left","top"))
popViewport()          
print(ggplot(star.logs[grepl("input reads$",field),],
             aes(x=capfirst(species),y=as.numeric(value)))
      + geom_boxplot()
      + scale_y_log10()
      + ylab("Reads")
      + xlab("Species")
      + theme(legend.position="top",
              axis.text.x = element_text(angle = 30, hjust = 1, vjust=1,face="italic")),
      vp=viewport(layout.pos.col=1,layout.pos.row=2))
pushViewport(viewport(layout.pos.col=1,layout.pos.row=2))
grid.text("B",
          y=unit(1,"npc")-unit(0.1,"cm"),
          x=unit(0,"npc")+unit(0.1,"cm"),
          gp=gpar(fontsize=24),
          just=c("left","top"))
popViewport(2)
@ 
\caption{RNAseq read and alignment statistics for all species. A)
  Boxplot of percentages of unmapped reads in each species; 0\% of
  reads in all species were unmapped because of mismatches. B) Boxplot
  of number of RNAseq reads in each species.}
\label{fig:transcriptome_statistics}
\end{figure}
